---
title: "tricycle_analysis Neuro subset"
author: "Jared Slosberg"
date: "4/5/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(monocle3)
library(tricycle)
library(tidyverse)
library(ggplot2)
library(here)
library(cowplot)

source(here("scripts/accessory_functions/monocle_mods.R"))
```

```{r}
name <- "tc_lmmp_p20"

lmmp <- readRDS(here(paste0(name,".rds")))
neuro <- lmmp[,pData(lmmp)$figure_celltype %in% c("MENs", "NENs", "NC-glia")]
# 
# #remove tricycle info from all age neuro subset
# rm_ind <- colnames(pData(neuro)) %in% c("tricycleRadius", "tricyclePosition")
# pData(neuro)[,rm_ind] <- NULL
# 
# rownames(neuro) <- fData(neuro)[,"gene_id_trimmed"]
# counts_list <- assays(neuro)
# 
# assay(neuro, "logcounts") <- normalized_counts(neuro)
# 
# plot_cells(neuro)
# 
# #Change IPAN to NENs for annotation by group
# pData(neuro)$cell_type_aggregate <- pData(neuro)$cell_type %>% str_replace(., "neuroglia/neuroendocrine", "NENs")
  
```
Regenerate UMAP embedding after removing p10 and p60 timepoints
```{r}
# set.seed(235)
# 
# neuro <- preprocess_cds(neuro, num_dim = 15)
# neuro <- reduce_dimension(neuro, preprocess_method = "PCA")
# plot_cells(neuro)
```



```{r, eval = F}


neuro <- tricycle::project_cycle_space(neuro, exprs_values = "logcounts")

#plot pca
p1 <- scater::plotReducedDim(neuro, dimred = "tricycleEmbedding") +
  labs(x = "Projected PC1", y = "Projected PC2") +
  ggtitle(sprintf("Projected cell cycle space (n=%d)",
                  ncol(neuro))) + 
  theme_bw(base_size = 14)

neuro <- estimate_cycle_position(neuro)

#plot pca
p2 <- scater::plotReducedDim(neuro, dimred = "tricycleEmbedding", colour_by = "tricyclePosition") +
  labs(x = "Projected PC1", y = "Projected PC2") +
  ggtitle(sprintf("Projected cell cycle space (n=%d)",
                  ncol(neuro))) + 
  theme_bw(base_size = 14)

#schwabe
neuro <- estimate_Schwabe_stage(neuro, gname.type = 'ENSEMBL', species = 'mouse', exprs_values = "logcounts")

scater::plotReducedDim(neuro, dimred = "tricycleEmbedding", colour_by = "CCStage") +
  labs(x = "Projected PC1", y = "Projected PC2") +
  ggtitle(sprintf("Projected cell cycle space (n=%d)",
                  ncol(neuro))) +
  theme_bw(base_size = 14)

##cylic scale
p <- plot_emb_circle_scale(neuro, dimred = 1, point.size = 3.5, point.alpha = 0.9) +
  theme_bw(base_size = 14)
legend <- circle_scale_legend(text.size = 5, alpha = 0.9)
p3 <- plot_grid(p, legend, ncol = 2, rel_widths = c(1, 0.4))

pca_emb_pos <- reducedDim(neuro, "tricycleEmbedding")
tric_pos <- as.data.frame(colData(neuro))[,"tricyclePosition"]

#Add back to cds object
reducedDim(neuro, "tricycleEmbedding") <- pca_emb_pos
pData(neuro)[,"tricyclePosition"] <- tric_pos

p<-plot_emb_circle_scale(neuro, dimred = "UMAP", color_by = "tricyclePosition")
p4 <- plot_grid(p, legend, ncol = 2, rel_widths = c(1, 0.1))

print(p1)
print(p2)
print(p3)
print(p4)



```
```{r, eval = F}
mens <- neuro[, pData(neuro)$cell_type == "MENs"]

plot_cells(mens)

#plot pca
p1 <- scater::plotReducedDim(mens, dimred = "tricycleEmbedding") +
  labs(x = "Projected PC1", y = "Projected PC2") +
  ggtitle(sprintf("Projected cell cycle space (n=%d)",
                  ncol(mens))) + 
  theme_bw(base_size = 14)



#plot pca
p2 <- scater::plotReducedDim(mens, dimred = "tricycleEmbedding", colour_by = "tricyclePosition") +
  labs(x = "Projected PC1", y = "Projected PC2") +
  ggtitle(sprintf("Projected cell cycle space (n=%d)",
                  ncol(mens))) + 
  theme_bw(base_size = 14)


scater::plotReducedDim(mens, dimred = "tricycleEmbedding", colour_by = "CCStage") +
  labs(x = "Projected PC1", y = "Projected PC2") +
  ggtitle(sprintf("Projected cell cycle space (n=%d)",
                  ncol(mens))) +
  theme_bw(base_size = 14)

##cylic scale
p <- plot_emb_circle_scale(mens, dimred = 1, point.size = 3.5, point.alpha = 0.9) +
  theme_bw(base_size = 14)
legend <- circle_scale_legend(text.size = 5, alpha = 0.9)
p3 <- plot_grid(p, legend, ncol = 2, rel_widths = c(1, 0.4))

pca_emb_pos <- reducedDim(mens, "tricycleEmbedding")
tric_pos <- as.data.frame(colData(mens))[,"tricyclePosition"]

p<-plot_emb_circle_scale(mens, dimred = "UMAP", color_by = "tricyclePosition")
p4 <- plot_grid(p, legend, ncol = 2, rel_widths = c(1, 0.1))

print(p1)
print(p2)
print(p3)
print(p4)
```

```{r, eval = F}
plot_cells_mod(neuro, reduction_method = "tricycleEmbedding", color_cells_by = "figure_celltype", cell_size = 0.75) + facet_wrap(~figure_celltype)
```
Confidence in cell cycle position estimates
```{r}
cart2pol <- function(x, y)
{
  r <- sqrt(x^2 + y^2)
  t <- atan(y/x)

  data.frame(tricycleRadius = r, theta = t)
}

radius <- reducedDim(neuro, "tricycleEmbedding")

pol_coord <- cart2pol(radius[,"PC1"], radius[,"PC2"]) %>% rownames_to_column(var = "cell_id")

pData(neuro) <- pData(neuro) %>% as.data.frame() %>% rownames_to_column(var = "cell_id") %>%
  left_join(., pol_coord) %>% column_to_rownames(var = "cell_id") %>% DataFrame()

ggplot(as.data.frame(pData(neuro)), aes(x = tricyclePosition, y = tricycleRadius)) + geom_point(aes(color = figure_celltype), size = .75) + facet_wrap(~figure_celltype) + geom_smooth()
```

Quantification of cycling cells, defined as 0.5pi < theta < 1.5pi

```{r}
cts <- pData(neuro) %>% as.data.frame() %>% pull(figure_celltype) %>% unique()
names(cts) <- cts
cycling_table <- map_df(cts, function(ct){
  
  sub <- neuro[,pData(neuro)$figure_celltype == ct]
  
  n_cycling <- sum(pData(sub)$tricyclePosition > 0.5*pi & pData(sub)$tricyclePosition < 1.5*pi) 
  
  prop_cycling <- n_cycling/dim(sub)[2]
  
  total = dim(sub)[2]
  
  data.frame(total, n_cycling, prop_cycling)

})

rownames(cycling_table) <- names(cts)

cycling_table
```


```{r}

#Continuous
tricycle::plot_ccposition_den(neuro$tricyclePosition,neuro$figure_celltype,
                              'cell type', type = "linear",bw = 10,
                              fig.title = "Kernel density of \u03b8") +
  theme_bw(base_size = 14)

tricycle::plot_ccposition_den(neuro$tricyclePosition,neuro$figure_celltype,
                              'cell type', type = "circular",bw = 10,
                              fig.title = "Kernel density of \u03b8") +
  theme_bw(base_size = 14)


#Discrete
as.data.frame.matrix(table(as.data.frame(pData(neuro)[,c("figure_celltype", "CCStage")]), useNA = "ifany"))
```
Assess gene expression over cell cycle
```{r, warning= F}
genes <- c("Top2a", "Pcna", "Ect2", "Ret", "Met")
pData(cds)$theta <- NULL

pls <- map(genes, function(gene){
  idx <- which(rowData(cds)$gene_short_name == gene)
  fit.l <- fit_periodic_loess_mod(cds, "tricyclePosition",
                              'logcounts', idx,
                              plot = TRUE, plot_loess = F,
                              color.by = "log10UMI",
                              point.size = 5,
                         x_lab = "Cell cycle position \u03b8", y_lab = "log2(Top2a)",
                         fig.title = glue::glue("Expression of ",gene, " along \u03b8 (n=", ncol(cds), ")"))
  names(fit.l)
  #> [1] "fitted"   "residual" "pred.df"  "loess.o"  "rsquared" "fig"
  fit.l$fig + theme_bw(base_size = 14) + facet_wrap(~cell_type) + geom_smooth(aes(x = theta, y = y)) + ylab("log10(exprs)") + viridis::scale_color_viridis()
})

cairo_pdf(here("plots/6mo_manuscript/gene_expression_over_theta.pdf"), height = 24, width = 18)
do.call(ggpubr::ggarrange, c(pls, ncol = 1))
dev.off()
```


```{r}
write.csv(coef, here("results/tricycle_cc_varying_genes_neuro.csv"))
write.csv(cycling_table, here("results/supp_figures/cycling_proportions_neuro_p20.csv"))

saveRDS(neuro, here("tc_neuro_p20.rds"))
```

