---
title: "6mo_MENS_followup.Rmd"
author: "Jared Slosberg"
date: "11/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r initialize, echo = F, message = F}
library(monocle3)
library(xtable)
library(scales)
library(pheatmap)
library(dplyr)
library(scran)
library(ggplot2)
library(org.Mm.eg.db)
library(ggpubr)
library(dplyr)
library(ComplexHeatmap)
library(NNLM)
library(viridis)
library(htmlwidgets)
library(here)
library(glue)

source(here("scripts/accessory_functions/monocle_mods.R"))
source(here("scripts/accessory_functions/doDifferentialExpression.R"))
source(here("scripts/accessory_functions/pattern_plotting.R"))
source(here("scripts/accessory_functions/geneSetEnrichment.R"))
source(here("scripts/accessory_functions/cluster_genes_mod.R"))
source(here("scripts/accessory_functions/transferAnnotations.R"))

set.seed(138)
```

```{r}
#MENS subset
lmmp <- readRDS(here("6month_LMMP.rds"))

#After running the above script and determining which quality cells are in the MEN population, subset on these


MENS <- lmmp[,pData(lmmp)$cell_type == "MENs"]

MENS <- estimate_size_factors(MENS)
MENS <- preprocess_cds(MENS, num_dim = 10, verbose=TRUE)
MENS <- reduce_dimension(MENS,max_components=2,cores=4,verbose=TRUE)
plot_cells(MENS)
plot_cells(MENS, color_cells_by = "sample", label_cell_groups = F)

pData(MENS)$batch<-pData(MENS)$sample
pData(MENS)$log10UMI <- log10(pData(MENS)$Total_UMIs)


MENS <- align_cds(MENS, num_dim = 10, residual_model_formula_str = "~ batch + num_genes_expressed",alignment_k = 20,verbose=T)
MENS <- reduce_dimension(MENS,max_components=2,cores=4,verbose=TRUE)
MENS <- cluster_cells(MENS, k=20, cluster_method="leiden",num_iter=5, reduction_method="UMAP", partition_qval=0.05,verbose=TRUE,resolution=1e-03)
plot_cells(MENS)
plot_cells(MENS, color_cells_by = "sample", label_cell_groups = F)

pData(MENS)$clusters<-clusters(MENS)
#Assigned cluster is a proxy for cluster where you can manually consolidate / separate clusters of cells
pData(MENS)$assigned_cluster <- pData(MENS)$clusters





```

```{r markers, eval = T, cache = T}
marker_test_res <- top_markers(MENS, group_cells_by="cluster", cores=8,verbose=T)
top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(25, pseudo_R2) %>% arrange(desc(cell_group))

dt <- DT::datatable(top_specific_markers)

htmlwidgets::saveWidget(dt, file = here("results/MENS/MENS_subset_cluster_markers.html"))





#pData(MENS)$cell_type[clusters(MENS) == 1] <- "Unknown"
```



```{r}
#pdf("/home/jared/ENS/Timecourse_ENS/plots/MENS.pdf")

plot_cells(MENS, group_label_size = 3)
plot_cells(MENS, color_cells_by = "log10UMI")
plot_cells(MENS, color_cells_by = "cycle_phase", cell_size = 0.6, label_cell_groups = F)

plot_cells_mod(MENS, color_cells_by="sample", group_cells_by="cluster", label_cell_groups = F) +
  ggtitle("Sample") + facet_wrap(vars(sample))


p<-ggplot(as.data.frame(pData(MENS))) +
  geom_bar(aes(x=clusters,fill=sample),position="fill") + scale_color_brewer() + monocle3:::monocle_theme_opts() + ggtitle("sample representation by cluster")
p1<-ggplot(as.data.frame(pData(MENS))) +
  geom_bar(aes(x=clusters,fill=batch),position="fill") + scale_color_brewer() + monocle3:::monocle_theme_opts() + ggtitle("batch representation by cluster")

p
p1
```


```{r, pseudotime, eval = T}
MENS <- learn_graph(MENS, verbose = TRUE)
MENS <- order_cells(MENS)

genes_of_interest <- c("Apoh","Msln")
indices <- as.integer(NULL)
for(gene in genes_of_interest){
 ind <- which(fData(MENS)$gene_short_name == gene)
 indices <- c(indices,ind)
}
plot_genes_in_pseudotime(MENS[indices,])
  

#MENS <- learn_graph(MENS, learn_graph_control = list("minimal_branch_len" = 2))
#MENS <- order_cells(MENS) #shiny app pop up



#MENS_sub <- MENS[,clusters(MENS)==2]
#MENS_sub <- learn_graph(MENS_sub,learn_graph_control = list("minimal_branch_len" = 2))

#plot_cells(MENS_sub, color_cells_by = "pseudotime")

goi <- c("Msln","Cftr","Fmo2")
plot_genes_in_pseudotime(MENS[fData(MENS)$gene_short_name %in% goi,], color_cells_by = "pseudotime", min_expr = 0.5)

MENS <- detect_genes(MENS)
res <- graph_test(MENS, neighbor_graph = "principal_graph")

res <- res %>% arrange(q_value)
res <- res[res$num_cells_expressed > 20,]
top_ids <- res$gene_short_name[1:20]


#p <- plot_genes_in_pseudotime(MENS[fData(MENS)$gene_short_name %in% top_ids[1:20],], color_cells_by = "pseudotime", min_expr = 0.5, panel_order = top_ids[1:20])
#ggsave("~/Goff_Lab/Kulkarni_ENS/Analysis/Plots/MENS_followup/trajectory_genes_pseudotime.pdf", height = 40)


```

```{r diff exp, eval = T}

lrt <- doDifferentialExpression(cds_obj = MENS, method = "LRT",min_pct_cells_exprs =  .02, 
                                full = "~num_genes_expressed + mito_ratio + clusters",
                                reduced = "~num_genes_expressed + mito_ratio")

write.csv(lrt, file = here("results/MENS/de_genes_by_cluster.csv"))
lrt_filt <- lrt %>% filter(q_value < 0.05)
print(qplot(lrt_filt$num_cells_expressed))
#ensembl ids character vector
de_genes <- pull(lrt_filt, gene_id)

# graph_test_res <- graph_test(MENS, neighbor_graph = "knn", k = 15, verbose = T)
# pData(cds)$test_group <- NULL
# pData(cds)$test_group[pData(cds)$assigned_cluster == 3] <- 3
# pData(cds)$test_group[pData(cds)$assigned_cluster == 2] <- 2
# plot_cells(cds, color_cells_by = "test_group")


```
##There are 'r length(de_genes)' differentially expressed genes by MENs subcluster (p < 0.05)


```{r, eval = T}

plot_cells(MENS, color_cells_by = "cluster")
plot_cells(MENS, color_cells_by = "CCStage",label_cell_groups = F)
```


```{r patternAnalysis, message=F}

set.seed(155)
date = "10-4"

MENS <- estimate_size_factors(MENS)
print(dim(MENS))

#use either (size -> log) normalized counts
#normed_exprs <- normalized_counts(MENS, norm_method = "log")
#or use just log normalized
normed_exprs <- normalized_counts(MENS, norm_method = "log") #and size normalized

#Choose dimensionality of NMF, how many patterns will be identified
npattern <- 40
system.time(decomp <- nnmf(A=as.matrix(normed_exprs),k = npattern,verbose = F, show.warning = T))
rownames(decomp$H) = paste0("cellPattern",c(1:npattern)) 

#Change this to something like cluster, cell type... whatever cell metadata to group by
group_feature <- "clusters"

#Subset to cell name and feature, group of feature to establish column rows
cells_by_group <- as.integer(pData(MENS)[,group_feature])
names(cells_by_group) <- colnames(MENS)
#order cells by their cluster, this object is a named integer array
cells_by_group <- cells_by_group[order(cells_by_group)]
cells_order <- names(cells_by_group)


#Plot heatmap of H (cell weight) matrix
#Assign a color for each group, these match default ggplot colors
my_colors <- scales::hue_pal()(length(levels(pData(MENS)[,group_feature])))
names(my_colors) <- unique(cells_by_group)
ha_cellgroup <- HeatmapAnnotation(cluster = cells_by_group,
                                  col = list(cluster = my_colors))

#We want cells grouped by cluster, as above. We also want to do hierarchical clustering
#on the pre-defined cluster level. Info must be in the same order 
cell_mat <- decomp$H[,cells_order]
stopifnot(colnames(cell_mat) == cells_order)


pdf(here(glue("plots/NMF/MENS/MENS_{date}_k_40_log_cellsize_cellweights.pdf")), width = 10)
plot_cells(MENS, color_cells_by = group_feature, group_label_size = 3)

hm <- ComplexHeatmap::Heatmap(cell_mat, name = "cellscore",
                        top_annotation = ha_cellgroup,
                        column_order = cells_order,
                        #cluster_columns = cluster_within_group(cell_mat, groups_to_cluster),
                        show_column_names = F)
                        #row_order = rownames(decomp$H))

hm
dev.off()

hm

##Gene weights associated with patterns
geneWeights.df <- as.data.frame(decomp$W)
colnames(geneWeights.df)<-paste0("cellPattern",c(1:npattern))
#Merge cell weights with pData for plotting pattern scores on UMAP embedding
cellWeights.df <- base::as.data.frame(t(decomp$H))
colnames(cellWeights.df) = paste0("cellPattern",c(1:npattern))

```


```{r}
#This adds pattern cell weights to pData, make sure they align
#left_join() would be more specific than this cbind()
add.to.cds <- 1
if(add.to.cds){
  stopifnot(rownames(pData(MENS)) == rownames(cellWeights.df))
  pData(MENS) <- cbind(pData(MENS),cellWeights.df)
}


#Color UMAP embedding by cell weights for each pattern.
#Call function and return a list of ggplot objects, and plot to one page
weighted_emb <- lapply(1:npattern,
                       FUN = plotCellPatterns,
                       cds_obj = MENS,
                       red_method <- "UMAP",
                       do.clip <- c(0.02,.98))

png(paste0("../plots/NMF/MENS/MENS_",date,"_NMF_Patterns.png"), height= 2000, width = 2000)
print(do.call(ggarrange, weighted_emb[1:npattern]))
dev.off()


tmp<-as.data.frame(merge(fData(MENS)[,c("gene_id","gene_short_name")],geneWeights.df,by=0))
dt<- DT::datatable(tmp[,c("gene_id","gene_short_name",
                      unlist(lapply(1:npattern,function(i){paste0("cellPattern",i)}))
)])

saveWidget(dt, file = here("results/NMF/MENS/MENS_10-31_pattern_gene_weights.html"))

write.csv(cellWeights.df, here(glue("results/NMF/MENS/MENS_{date}_log_cellsize_pattern_cell_weights.csv")))

write.csv(geneWeights.df, here(glue("results/NMF/MENS/MENS_{date}_log_cellsize_pattern_gene_weights.csv")))

```

```{r PatternCorr}

#### Check correlation of patterns with cell features
#### One hot encoding method.
### For each pattern, compute correlation between two vectors. One is continuous (cell pattern weight)
### and one is binary (yes/no is that cell in that group)
pData(MENS)$cycle_phase[is.na(pData(MENS)$cycle_phase)] <- "NA"

#categorical features, one-hot encoding
features <- c("clusters","cycle_phase")
heatmap_figure_list <- lapply(features, function(feature){
  feature_levels <- sort(unique(pData(MENS)[,feature]))
  
  corr_mat <- matrix(nrow = length(feature_levels), ncol = npattern)
  
  
  #loop over each pattern
  for(i in 1:npattern){
    cell_weights <- cellWeights.df[,paste0("cellPattern",i)]
    names(cell_weights) <- rownames(cellWeights.df)
    #re-initialize empty correlation vector
    corr_vec <- numeric()
    #loop over each level within the categorical variable
    for(level in feature_levels){
  
      values <- pData(MENS)[,feature]
      one_hot_values <- (values == level)
      names(one_hot_values) <- colnames(MENS)
      #Make sure our cells are in the same order
      assertthat::assert_that(sum(names(one_hot_values) == names(cell_weights)) == length(one_hot_values))
      
      #calculate correlation between two vectors
      level_cor <- cor(cell_weights, one_hot_values)
      #corr_df is a one-dimensional vector, eg the correlations between all features for a single pattern
      corr_vec <- c(corr_vec, level_cor)
    }
    #Now we have a n_feature_levels x n_pattern matrix with pairwise correlations
    corr_mat[,i] <- corr_vec
  }
  rownames(corr_mat) <- feature_levels
  colnames(corr_mat) <- paste0("cellPattern",1:npattern)
  
  heatmap_figure <- ComplexHeatmap::Heatmap(corr_mat, name = "pearson_corr", 
                                            column_title = paste0("correlation between pattern and ",feature, " annotation"))
  return(heatmap_figure)
})


pdf(here(glue("plots/NMF/MENS/MENS_{date}_pattern_11-25_categorical_feature_correlation.pdf")), height= 8, width = 8)
print(heatmap_figure_list)
dev.off()




```

```{r GSEA, eval = F, message=F, warning= F}

geneWeights.df <- geneWeights.df %>% tibble::rownames_to_column(var = "gene_id")
geneSetEnrichment(gene_weights = geneWeights.df, n_genes = 1500, n_patterns = 40,
                  file_prefix = here(glue("results/GSEA/mens/MENS_{date}_pattern_gene_weights"))

```

##Usage of patterns learned in LMMP 
```{r}
#Remove existing pattern weights
pData(MENS) <- dplyr::select(as.data.frame(pData(MENS)), !starts_with("cellPattern")) %>% DataFrame()

#load and name column 1 with cell ids. can't be rownames for merge
lmmp_cell_wts <- read.csv(here("results/NMF/lmmp/old_pattern_run/50dims/pattern_cell_weights.csv"))
colnames(lmmp_cell_wts)[1] <- "cell_id"

lmmp_gene_wts <- read.csv(here("results/NMF/lmmp/old_pattern_run/50dims/pattern_gene_weights.csv"), row.names = 1)

n_lmmp_patts <- 50

pdat <- pData(MENS) %>% as.data.frame() %>% mutate("cell_id" = paste(barcode, sample, sep = "."))

pData(MENS) <- left_join(pdat,lmmp_cell_wts, by= "cell_id") %>% DataFrame()
rownames(pData(MENS)) <- pData(MENS)$cell_id

```



```{r}
#Color UMAP embedding by cell weights for each pattern.
#Call function and return a list of ggplot objects, and plot to one page
weighted_emb <- lapply(1:n_lmmp_patts,
                       FUN = plotCellPatterns,
                       cds_obj = MENS,
                       red_method <- "UMAP",
                       do.clip <- c(0.02,.98))

png(here("plots/NMF/MENS/MENS_subset_lmmp_patterns.png"), height= 2000, width = 2000)
print(do.call(ggarrange, weighted_emb[1:n_lmmp_patts]))
dev.off()

```
 


```{r}
saveRDS(MENS, file = here(glue("6mo_MENS.rds")))
sessionInfo()
```
