---
title: "6mo LMMP pattern identification"
author: "Jared Slosberg"
date: "9/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = F, results="hide"}
library(monocle3)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(ComplexHeatmap)
library(htmlwidgets)
library(NNLM)
library(here)
library(purrr)
library(glue)
source(here("scripts/accessory_functions/pattern_plotting.R"))
source(here("scripts/accessory_functions/monocle_mods.R"))
source(here("scripts/accessory_functions/geneSetEnrichment.R"))
source(here("scripts/accessory_functions/patternFeatureCorrelationHeatmap.R"))

```


```{r}
lmmp_6mo <- readRDS(here("6month_LMMP.rds"))
set.seed(42)

lmmp_6mo <- estimate_size_factors(lmmp_6mo)
print(dim(lmmp_6mo))

date <- "10-7"

#Change this to something like cluster, cell type... whatever cell metadata to group by
group_feature <- "clusters"
```

## Generate patterns, or load in existing patterns
```{r}
do.existing.patterns <- 1
existing_pattern_path <- here("results/NMF/lmmp/old_pattern_run/50dims/")
if(do.existing.patterns){
  date <- ""
  npattern <- 50
  geneWeights.df <- read.csv(paste0(existing_pattern_path, "pattern_gene_weights.csv"), row.names = 1)
  cellWeights.df <- read.csv(paste0(existing_pattern_path, "pattern_cell_weights.csv"), row.names = 1)
} else{

  #use either (size -> log) normalized counts
  #normed_exprs <- normalized_counts(lmmp_6mo, norm_method = "log")
  #or use just log normalized
  normed_exprs <- log10(as.matrix(exprs(lmmp_6mo))+1)
  
  #Choose dimensionality of NMF, how many patterns will be identified
  npattern <- 50
  system.time(decomp <- nnmf(A=as.matrix(normed_exprs),k = npattern,verbose = F, show.warning = T))
  rownames(decomp$H) = paste0("cellPattern",c(1:npattern)) 
  
  ##Gene weights associated with patterns
  geneWeights.df <- as.data.frame(decomp$W)
  colnames(geneWeights.df)<-paste0("cellPattern",c(1:npattern))
  #Merge cell weights with pData for plotting pattern scores on UMAP embedding
  cellWeights.df <- base::as.data.frame(t(decomp$H))
  colnames(cellWeights.df) = paste0("cellPattern",c(1:npattern))
}

#Subset to cell name and feature, group of feature to establish column rows
  cells_by_group <- as.integer(pData(lmmp_6mo)[,group_feature])
  names(cells_by_group) <- colnames(lmmp_6mo)
  #order cells by their cluster, this object is a named integer array
  cells_by_group <- cells_by_group[order(cells_by_group)]
  cells_order <- names(cells_by_group)
  
  
  #Plot heatmap of H (cell weight) matrix
  #Assign a color for each group, these match default ggplot colors
  my_colors <- scales::hue_pal()(length(levels(pData(lmmp_6mo)[,group_feature])))
  names(my_colors) <- unique(cells_by_group)
  ha_cellgroup <- HeatmapAnnotation(cluster = cells_by_group,
                                    col = list(cluster = my_colors))
  
  #We want cells grouped by cluster, as above. We also want to do hierarchical clustering
  #on the pre-defined cluster level. Info must be in the same order 
  cell_mat <- t(cellWeights.df[cells_order,])
  stopifnot(colnames(cell_mat) == cells_order)


```

## Visualize pattern usage by heatmap
```{r}
plot_cells(lmmp_6mo, color_cells_by = group_feature, group_label_size = 3)

ComplexHeatmap::Heatmap(cell_mat, name = "cellscore",
                        top_annotation = ha_cellgroup,
                        column_order = cells_order,
                        #cluster_columns = cluster_within_group(cell_mat, groups_to_cluster),
                        show_column_names = F)
#row_order = rownames(decomp$H))
```

## Add the pattern weights to pData
```{r}
#This adds pattern cell weights to pData, make sure they align
#left_join() would be more specific than this cbind()
add.to.cds <- 1
if(add.to.cds){
  stopifnot(rownames(pData(lmmp_6mo)) == rownames(cellWeights.df))
  pData(lmmp_6mo) <- cbind(pData(lmmp_6mo),cellWeights.df)
}


```

## Pattern usage on UMAP
```{r}
#Color UMAP embedding by cell weights for each pattern.
#Call function and return a list of ggplot objects, and plot to one page
weighted_emb <- lapply(1:npattern,
                       FUN = plotCellPatterns,
                       cds_obj = lmmp_6mo,
                       red_method = "UMAP",
                       do.clip = c(0.02,.98))

print(weighted_emb[1:npattern])

```

## Pattern usage by sample
```{r}
#Look at pattern usage over sample

condition_patterns<- as.list(1:npattern)
myplots <- lapply(paste0("cellPattern",condition_patterns), 
                  FUN= plotPatternUsageByCondition,
                  cds = lmmp_6mo,
                  bin_by = "sample")
n_by <- 10
seq_by <- seq(0, npattern, n_by)
map(1:(length(seq_by)-1), function(i){
  print(i)
  print(do.call(ggarrange, myplots[(seq_by[i]+1):(seq_by[i+1])]))
})


```

## Save results
```{r}

if(!do.existing.patterns){
  tmp<-as.data.frame(merge(fData(lmmp_6mo)[,c("gene_id","gene_short_name")],geneWeights.df,by=0))
  dt<- DT::datatable(tmp[,c("gene_id","gene_short_name",
                       unlist(lapply(1:npattern,function(i){paste0("cellPattern",i)}))
  )])
  
  saveWidget(dt, file = paste0("/home/jared/ENS/6mo_LMMP/results/NMF/lmmp/lmmp_6mo_",date,"_pattern_gene_weights.html"))
  
  write.csv(cellWeights.df, paste0("/home/jared/ENS/6mo_LMMP/results/NMF/lmmp/lmmp_6mo_",date,"_pattern_cell_weights.csv"))
  
  write.csv(geneWeights.df, paste0("/home/jared/ENS/6mo_LMMP/results/NMF/lmmp/lmmp_6mo_",date,"_pattern_gene_weights.csv"))
}
```

## Run gene set enrichment analysis
```{r}

if(!do.existing.patterns){
  #Run GSEA 
  geneWeights.df <- geneWeights.df %>% tibble::rownames_to_column(var = "gene_id")
  
    #putting gene names as rownames "cleaned" them, so genes starting with number or having "-" were changed. 
  geneWeights.df.filt <- geneWeights.df
  geneWeights.df.filt[,"gene_id"] <- rownames(lmmp_6mo)
  
  geneSetEnrichment(gene_weights = geneWeights.df.filt,
                    n_genes = 1000,
                    n_patterns = 50,
                    file_prefix =
                      paste0(here("results/GSEA/lmmp/lmmp_6mo_",date,"_"))
                    )
}
```

## Look at pattern correlations with metadata
```{r}
#Run pattern correlations
disc_features <- c("cell_type","CCStage", "sample")
cont_features <- c("log10UMI","mito_ratio","tricyclePosition")

pData(lmmp_6mo)$CCStage[is.na(pData(lmmp_6mo)$CCStage)] <- "NA"

heatmaps <- patternFeatureCorrelationHeatmap(lmmp_6mo, cellWeights.df = cellWeights.df, discrete_features = disc_features, continuous_features = cont_features)

hm_edit <- lapply(heatmaps[[1]], function(x){
  x@column_names_param$gp$fontsize <- 8
  x
})

if(!do.existing.patterns){
  
  pdf(here(glue("plots/NMF/lmmp/LMMP_{date}_pattern_feature_correlation.pdf")), width= 12)
  hm_edit[[1]]@row_names_param$gp$fontsize <- 10
  hm_edit
  
  
  
  Heatmap(heatmaps[[2]], column_title = "Correlation of pattern weights with continuous features",
          name = "Pearson",
          column_names_gp = gpar(fontsize = 8),
          column_names_rot = 70,
          row_names_gp = gpar(fontsize = 10))
  
  dev.off()
}

```

# Dont save cds object because pattern names are ambiguous between runs. In the future load the patterns explicitly everytime they are used.